\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A Radial Time Projection Chamber for $\alpha$ detection in CLAS at JLab}

### 2017-06-06 eg6-letters/NIM/EG6_RTPC_NIM.tex
add the reference for figure 3
Check typos

Thomas Jefferson National Accelerator Facility (TJNAF), in Newport News, Virginia,
USA, provides high power electron beams of energy up to 6 GeV and 100$\%$ duty cycle
to three experimental Halls (A, B, C) simultaneously. The CLAS 
spectrometer~\cite{CLASref}, located
in Hall-B, is composed of several sub-detectors and two magnets. 
Figure~\ref{fig:CLAS} shows a three dimensional 
representation of the baseline CLAS spectrometer:
\begin{itemize}
 \item Three regions of Drift Chambers (DC) for the tracking of charged 
    particles~\cite{DCref} (in blue on the figure).
 \item Superconducting toroidal magnet (in yellow) to bend the trajectories 
    of charged particles, thus allowing momentum measurement with the DC tracking information.
 \item Cherenvkov Counters (CC) to distinguish electrons from other negative 
    particles, such as pions, at lower energies~\cite{CCref} (in pink).
 \item Scintillation Counters (SC) to identify hadrons by measuring their 
    time of flight (TOF)~\cite{TOFref} (in red).
 \item Electromagnetic Calorimeters (EC) to distinguish electrons from other negative 
    particles at higher energies and to detect neutral particles (neutrons and 
    photons in particular)~\cite{ECref} (in green).
\end{itemize}

@@

The GEM technology has been chosenfor its flexibility as it
allows for a curved amplification surface. Also, GEMs are known to rarely create 
sparks~\cite{GEM_ref}, which is important when trying to detect 
highly ionizing slow nuclei that deposit large amount of energy. The GEMs for 
this RTPC are made from a Kapton insulator layer, 50~$\mu$m 
thick, sandwiched between two 5~$\mu$m copper layers. The mesh of each GEM 
layer is chemically etched with 50~$\mu$m diameter holes with double-conical 
shapes as illustrated in figure~\ref{fig:GEMs}. The potential difference 
applied between the two copper layers of the GEM creates a very strong 
electric field in each hole leading to high ionization and amplification.

The drift gas used in the experiment is a 80-20\% neon-DME mixture, this choice 
has been made in order to balance the energy deposit with a reasonnable
Lorentz angle. Calculations using the MAGBOLTZ program~\cite{MAGBOLTZ} 
showed that with the perpendicular 5~T field, we would have a Lorentz angle of 
about 23$^\circ$ with this gas mixture. This is already a significant angle,
a heavier gas would have larger angle increasing significantly
the length of the drift paths as the electron would 
make a long spiral around the beam line before to reach the GEMs.

One of the important problematic to develop the radial TPC was to obtain a 
good support structure of the GEM foils, allowing an easy installation of the 
GEM in good conditions and with high precision. At the same time, we wanted to 
keep the material budget small in the forward region where we detect other 
particles in subsequent detectors. We successfuly realized these prerequisites 
by using fiber glass rings glued to each ends of the GEM foils to form self 
supporting cylinders that could be installed independently in the RTPC after 
all gluing and soldering operations. The rigidity of the GEM foil is enough
for the structure to be selfsupporting and only the 
backward part of the cylinder was fixed to the main mechanical structure. This 
design only left a light fiberglass ring in the forward region, reducing to a 
minimum secondary interactions.

@@

The readout system is based on the Front End Electronic (FEE) boards originally 
developed for the ALICE TPC readout system at Cern~\cite{ALICE-FEE}. Each 
readout channel is made of three main components: a charge sensitive, low 
impedance shaping amplifier, a 10 bit, 25-MHz digitizer, and a digital circuit 
that implements online-processing algorithms (pedestal subtraction, zero 
suppression, ad tail cancellation). The amplifier is implemented in a 
fully-analogue ASIC, PASA~\cite{ALICE-PASA}, while the subsequent signal 
digitization and elaboration part is implemented in a digital ASIC, 
ALTRO~\cite{ALICE-ALTRO}. Both ASICs host 16 channels. The full readout 
chain is implemented on a single Front End Card (FEC), hosting 128 channels. 
FECs are hosted in custom crates, mounted close to the detector. A Readout 
Control Unit (ROC) board is used to distribute the trigger signal to FECs 
and for data readout. Each RCU can handle up to 25 FECs. Communication 
between ROC and FECs is performed trough a custom backplane, implementing 
a low-voltage signal bus. The RCU communicates with the CLAS DAQ system 
trough a 200 MB/s optical link, connected to a data acquisition PC hosting 
a ReadOut Receiver Card (RORC). This PC hosts the readout controller 
application (ROC) used to interface with the CLAS DAQ system. An Ethernet 
link is also present, for slow-controls and monitoring.

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. Software exists to calculate the drift paths, in particular
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used such calculation as a 
first calibration, but, 
as can be seen with the drift speed, gas composition is far from stable in the 
chamber. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometry accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.20 and 1.27 GeV) to enhance the cross section of the elastic 
scatering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematic allows to calculate the Helium nucleus kinematic. It is
by comparing this calculated kinematic to the measured one that we fine tune
the drift paths independently of our knowledge of the exact conditions in the 
chamber.

The drift paths are ajusted using a set of identified elastic events. We generate 
a helium nucleus in a GEANT4 simulation~\cite{GEANT4} of our RTPC for each event
based on the measured electron kinematic. Then, we compare the simulated trajectory of 
the Helium nuclei to the hits measured in the chamber. Because of the 
magnetic field, the drift paths are not linear in the RTPC. So to perform the extraction, 
we make a first approximation with a linear dependence between the radius of 
emission and the time of detection, and then refine our result. As it happens, 
the curvature is minimal and this process converge already on the second 
iteration.

@@

The second is a coherent noise affecting about 25\% of the pramplifiers, where the signature is simultaneous hits in most of the 16 channels in a preamp group.   An event-based technique to identify and remove it was developed based on counting simultaneous hits in preamp group, and, if sufficiently large, perform a dynamic pedestal subtraction based on the average ADC of the channels' neighbors within the preamp group.

@@

We reported on the construction, operation and calibration of a small GEM based RTPC 
designed to measure helium-4 nuclei in a high rate environnement. The operation
of the detector was successful and allowed to detect helium nuclei at a rate
of 3.1 kHz in coincidence with the detection of high energy electrons and 
photons in the CLAS spectrometer.

\input acknowledgement.tex

>>>>

Thomas Jefferson National Accelerator Facility (TJNAF), in Newport News, Virginia,
USA, provides high power electron beams of energy up to 6 GeV and 100$\%$ duty cycle
to three experimental Halls (A, B, C) simultaneously. The CLAS 
spectrometer~\cite{CLASref}, located
in Hall-B, is composed of several sub-detectors and two magnets. 
Figure~\ref{fig:CLAS} shows a three dimensional 
representation of the baseline CLAS spectrometer:
\begin{itemize}
 \item Three regions of Drift Chambers (DC) for the tracking of charged 
    particles~\cite{DCref} (in blue on the figure).
 \item Superconducting toroidal magnet (in yellow) to bend the trajectories 
    of charged particles, thus allowing momentum measurement with the DC tracking information.
 \item Cerenkov Counters (CC) to distinguish electrons from other negative 
    particles, such as pions, at lower energies~\cite{CCref} (in pink).
 \item Scintillation Counters (SC) to identify hadrons by measuring their 
    time of flight (TOF)~\cite{TOFref} (in red).
 \item Electromagnetic Calorimeters (EC) to distinguish electrons from other negative 
    particles at higher energies and to detect neutral particles (neutrons and 
    photons in particular)~\cite{ECref} (in green).
\end{itemize}

@@

The GEM technology has been chosen for its flexibility as it
allows for a curved amplification surface. Also, GEMs are known to rarely create 
sparks~\cite{GEM_ref}, which is important when trying to detect 
highly ionizing slow nuclei that deposit large amount of energy. The GEMs for 
this RTPC are made from a Kapton insulator layer, 50~$\mu$m 
thick, sandwiched between two 5~$\mu$m copper layers. The mesh of each GEM 
layer is chemically etched with 50~$\mu$m diameter holes with double-conical 
shapes as illustrated in figure~\ref{fig:GEMs}. The potential difference 
applied between the two copper layers of the GEM creates a very strong 
electric field in each hole leading to high ionization and amplification.

The drift gas used in the experiment is a 80-20\% neon-DME mixture, this choice 
has been made in order to balance the energy deposit with a reasonable
Lorentz angle. Calculations using the MAGBOLTZ program~\cite{MAGBOLTZ} 
showed that with the perpendicular 5~T field, we would have a Lorentz angle of 
about 23$^\circ$ with this gas mixture. This is already a significant angle,
a heavier gas would have larger angle increasing significantly
the length of the drift paths as the electron would 
make a long spiral around the beam line before to reach the GEMs.

One of the important problematic to develop the radial TPC was to obtain a 
good support structure of the GEM foils, allowing an easy installation of the 
GEM in good conditions and with high precision. At the same time, we wanted to 
keep the material budget small in the forward region where we detect other 
particles in subsequent detectors. We successfully realized these prerequisites 
by using fiber glass rings glued to each ends of the GEM foils to form self 
supporting cylinders that could be installed independently in the RTPC after 
all gluing and soldering operations. The rigidity of the GEM foil is enough
for the structure to be self supporting and only the 
backward part of the cylinder was fixed to the main mechanical structure. This 
design only left a light fiberglass ring in the forward region, reducing to a 
minimum secondary interactions.

@@

The readout system is based on the Front End Electronic (FEE) boards originally 
developed for the ALICE TPC readout system at CERN~\cite{ALICE-FEE}. Each 
readout channel is made of three main components: a charge sensitive, low 
impedance shaping amplifier, a 10 bit, 25-MHz digitizer, and a digital circuit 
that implements online-processing algorithms (pedestal subtraction, zero 
suppression, ad tail cancellation). The amplifier is implemented in a 
fully-analogue ASIC, PASA~\cite{ALICE-PASA}, while the subsequent signal 
digitization and elaboration part is implemented in a digital ASIC, 
ALTRO~\cite{ALICE-ALTRO}. Both ASICs host 16 channels. The full readout 
chain is implemented on a single Front End Card (FEC), hosting 128 channels. 
FECs are hosted in custom crates, mounted close to the detector. A Readout 
Control Unit (ROC) board is used to distribute the trigger signal to FECs 
and for data readout. Each RCU can handle up to 25 FECs. Communication 
between ROC and FECs is performed trough a custom back-plane, implementing 
a low-voltage signal bus. The RCU communicates with the CLAS DAQ system 
trough a 200 MB/s optical link, connected to a data acquisition PC hosting 
a ReadOut Receiver Card (RORC). This PC hosts the readout controller 
application (ROC) used to interface with the CLAS DAQ system. An Ethernet 
link is also present, for slow-controls and monitoring.

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. Software exists to calculate the drift paths, in particular
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used such calculation as a 
first calibration, but, 
as can be seen with the drift speed, gas composition is far from stable in the 
chamber. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometry accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.20 and 1.27 GeV) to enhance the cross section of the elastic 
scattering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematic allows to calculate the Helium nucleus kinematic. It is
by comparing this calculated kinematic to the measured one that we fine tune
the drift paths independently of our knowledge of the exact conditions in the 
chamber.

The drift paths are adjusted using a set of identified elastic events. We generate 
a helium nucleus in a GEANT4 simulation~\cite{GEANT4} of our RTPC for each event
based on the measured electron kinematic. Then, we compare the simulated trajectory of 
the Helium nuclei to the hits measured in the chamber. Because of the 
magnetic field, the drift paths are not linear in the RTPC. So to perform the extraction, 
we make a first approximation with a linear dependence between the radius of 
emission and the time of detection, and then refine our result. As it happens, 
the curvature is minimal and this process converge already on the second 
iteration.

@@

The second is a coherent noise affecting about 25\% of the pre-amplifiers, where the signature is simultaneous hits in most of the 16 channels in a preamp group.   An event-based technique to identify and remove it was developed based on counting simultaneous hits in preamp group, and, if sufficiently large, perform a dynamic pedestal subtraction based on the average ADC of the channels' neighbors within the preamp group.

@@

We reported on the construction, operation and calibration of a small GEM based RTPC 
designed to measure helium-4 nuclei in a high rate environment. The operation
of the detector was successful and allowed to detect helium nuclei at a rate
of 3.1 kHz in coincidence with the detection of high energy electrons and 
photons in the CLAS spectrometer.

\input acknowledgment.tex
